Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2016-01-17 cordillera-cycle.tex
Typeset: remove comment lines.
Typeset: more no-break space after 'a' and 'A'.

% =============================================================================
\begin{document}
% =============================================================================

@@

% =============================================================================
\introduction
\label{sec:intro}
% =============================================================================

@@

      Although numerical ice sheet modelling has been established as a
      useful tool to improve our understanding of the Cordilleran ice sheet
      (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991};
      \citealp{Marshall.etal.2000}), the ubiquitously mountainous topography
      of the region has presented two major challenges to its application.
      First, only recent developments in numerical ice sheet models and
      underlying scientific computing tools \citep{Bueler.Brown.2009,
      Balay.etal.2015} have allowed for high-resolution numerical modelling
      of glaciers and ice sheets on mountainous terrain over millennial time
      scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the complex
      topography of the North American Cordillera also induces strong
      geographic variations in temperature and precipitation, thus requiring
      the use of high-resolution climate forcing fields as an input to an
      ice sheet model \citep{Seguinot.etal.2014}. However, evolving climate
      conditions over the last glacial cycle are subject to considerable
      uncertainty and still lie beyond the computational reach of atmosphere
      circulation models.

@@

% =============================================================================
\section{Model setup}
\label{sec:model}
% =============================================================================

% -----------------------------------------------------------------------------
\subsection{Overview}
\label{sec:overview}
% -----------------------------------------------------------------------------

@@

      Basal topography is derived from the ETOPO1 combined topography and
      bathymetry dataset with a~resolution of 1\,arc-min
      \citep{Amante.Eakins.2009}. Sea level is lowered as a~function of time
      based on the Spectral Mapping Project
      \citep[SPECMAP,][]{Imbrie.etal.1989} time scale. Geothermal heat flux
      is applied as a~constant value of 70\,\unit{mW\,m^{-2}} at
      3\,\unit{km} depth (Sect.~\ref{sec:icedyn}). Surface mass balance is
      computed using a positive degree-day (PDD) model
      (Sect.~\ref{sec:surface}). Climate forcing is provided by a~monthly
      climatology averaged from 1979 to 2000 from the North American
      Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, perturbated
      by time-dependent offsets and lapse-rate temperature corrections
      (Sect.~\ref{sec:atm}).

@@

% -----------------------------------------------------------------------------
\subsection{Ice thermodynamics}
\label{sec:icedyn}
% -----------------------------------------------------------------------------

      Ice sheet dynamics are typically modelled using a~combination of
      internal deformation and basal sliding. PISM is a~shallow ice sheet
      model, which implies that the balance of stresses is approximated
      based on their predominant components. The Shallow Shelf Approximation
      (SSA) is used as a~``sliding law'' for the Shallow Ice Approximation
      (SIA) by adding velocity solutions of the two approximations
      \citep[Eqns.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}.

@@

      Ice shelf calving is computed using a~double criterion. First, a
      physically-realistic calving flux is computed based on eigenvalues of
      the horizontal strain rate tensor \citep{Winkelmann.etal.2011,
      Levermann.etal.2012}. This allows floating ice to advance in confined
      embayments, but prevents the formation of extensive ice shelves in the
      open ocean. Second, floating ice thinner than 50\,\unit{m} is
      systematically calved off. A~subgrid scheme by
      \citet{Albrecht.etal.2011} allows for a~continuous migration of the
      calving front. This formulation of calving has been applied to the
      Antarctic ice sheet and has shown to produce a~realistic calving front
      positon for many of the present-day ice shelves
      \citep{Martin.etal.2011}.

% -----------------------------------------------------------------------------
\subsection{Surface mass balance}
\label{sec:surface}
% -----------------------------------------------------------------------------

@@

% -----------------------------------------------------------------------------
\subsection{Climate forcing}
\label{sec:atm}
% -----------------------------------------------------------------------------

      Climate forcing driving ice sheet simulations consists of a
      present-day monthly climatology,
$\{T_{m0}, P_{m0}\}$, where temperatures are modified by offset time series,
${\Delta}T_{TS}$, and lapse-rate corrections, ${\Delta}T_{LR}$:
\begin{align}
    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t)
                    + {\Delta}T_{LR}(t, x, y) \,, \\
    P_m(t, x, y) &= P_{m0}(x, y) \,.
\end{align}

@@

      Finally, lapse-rate corrections, ${\Delta}T_{LR}$, are computed as a
      function of ice surface elevation, $s$, using the NARR surface
      geopotential height invariant field as a~reference topography,
      $b_{ref}$:
\begin{align}
    {\Delta}T_{LR}(t, x, y) &= -\gamma [s(t, x, y)-b_{ref}] \\
                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{ref}],
\end{align}

@@

% =============================================================================
\section{Sensitivity to climate forcing time-series}
\label{sec:results}
% =============================================================================

% -----------------------------------------------------------------------------
\subsection{Evolution of ice volume}
% -----------------------------------------------------------------------------

@@

% -----------------------------------------------------------------------------
\subsection{Extreme configurations}
% -----------------------------------------------------------------------------

@@

      In the MIS~3 ice volume minimum reconstructions, a~central ice cap
      persists over the Skeena Mountains (Fig.~\ref{fig:lr_maps}, middle
      panels). Although this ice cap is present in all simulations, its
      dimensions depend sensitively on the choice of the applied
      palaeo-temperature record. Modelled ice volume minima spread over a
      wide range between 1.69 and 2.88\,\unit{m}\,s.l.e.
      (Table~\ref{tab:extrema}).

@@

% =============================================================================
\section{Comparison with the geologic record}
\label{sec:discussion}
% =============================================================================

@@

% -----------------------------------------------------------------------------
\subsection{Glacial maxima}
% -----------------------------------------------------------------------------

@@

      The exact timing of modelled MIS~2 maximum ice volume depends strongly
      on the choice of applied palaeo-temperature record, which allows for a
      more in-depth comparison with geological evidence for the timing of
      maximum Cordilleran ice sheet extent. In the Puget Lowland
      (Fig.~\ref{fig:locmap}), the LGM advance of the southern Cordilleran
      ice sheet margin has been constrained by radiocarbon dating on wood
      between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,ka}
      \citep{Porter.Swanson.1998}. These dates are consistent with
      radiocarbon dates from the offshore sedimentary record, which reveals
      an increase of glaciomarine sedimentation between 19.5 and
      16.2\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008,
      Taylor.etal.2014}. Radiocarbon dating of the northern Cordilleran ice
      sheet margin is much less constrained but straddles presented
      constraints from the southern margin. However, cosmogenic exposure
      dating places the timing of maximum CIS extent during the McConnell
      glaciation close to 17\,\unit{\chem{^{10}Be}\,ka}
      \citep{Stroeven.etal.2010, Stroeven.etal.2014}. A~sharp transition in
      the sediment record of the Gulf of Alaska indicates a~retreat of
      regional outlet glaciers onto land at
      14.8\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Davies.etal.2011}.

@@

      During maximum glaciation, both simulations position the main
      meridional ice divide over the western flank of the Rocky Mountains
      (Fig.~\ref{fig:lr_maps}, lower panels; Fig.~\ref{fig:mis2}). This
      result appears to contrast with palaeoglaciological reconstructions
      for central and southern British Columbia with ice divides in a~more
      westerly position, over the western margin of the Interior Plateau
      \citep{Ryder.etal.1991, Stumpf.etal.2000, Kleman.etal.2010,
      Clague.Ward.2011, Margold.etal.2013a}. These indicate that a
      latitudinal saddle connected ice dispersal centres in the Columbia
      Mountains with the main ice divide \citep{Ryder.etal.1991,
      Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A~latitudinal
      saddle does indeed feature in our modelling results, however, in an
      inverse configuration between the main ice divide over the Columbia
      Mountains and a~secondary divide over the southern Coast Mountains
      (Fig.~\ref{fig:mis2}).

      Such deviation from the geological inferences could reflect the fact
      that our model does not include feedback mechanisms between ice sheet
      topography and the regional climate. Firstly, during the build-up
      phase preceding the LGM, rapid accumulation over the Coast Mountains
      enhanced the topographic barrier formed by these mountain ranges,
      which likely resulted in a~decrease of precipitation and, therefore, a
      decrease of accumulation in the interior. Secondly, latent warming of
      the moisture-depleted air parcels flowing over this enhanced
      topography could have resulted in an inflow of potentially warmer air
      over the eastern flank of the ice sheet, increasing melt along the
      advancing margin \citep[cf.][]{Langen.etal.2012}. Because these two
      processes, both with a~tendency to limit ice-sheet growth, are absent
      from our model, the eastern margin of the ice sheet and the position
      of the main meridional ice divide are certainly biased towards the
      east in our simulations \citep{Seguinot.etal.2014}.

@@

% -----------------------------------------------------------------------------
\subsection{Nucleation centres}
% -----------------------------------------------------------------------------

@@

      Perhaps the most striking feature displayed by the distributions of
      modelled ice cover is the persistence of the Skeena Mountains ice cap
      throughout the entire last glacial period (ca.~100--10\,\unit{ka}) and
      its predominance over the other ice-dispersal centres
      (Figs.~\ref{fig:lr_maps} and~\ref{fig:hr_geom_duration}, SM).
      Regardless of the applied forcing, this ice cap appears to survive
      MIS~3 (Fig.~\ref{fig:lr_maps}, middle panels), and serves as a
      nucleation centre at the onset of the glacial readvance towards the
      LGM (MIS~2). This situation appear similar to the neighbouring
      Laurentide ice sheet, for which the importance of residual ice for the
      glacial history leading up to the LGM has been illustrated by the
      MIS~3 residual ice bodies in northern and eastern Canada as nucleation
      centres for a~much more extensive MIS~2 configuration
      \citep{Kleman.etal.2010}.

      The presence of a~Skeena Mountains ice cap during most of the last
      glacial cycle can be explained by meteorological conditions more
      favourable for ice growth there than elsewhere. In fact, reanalysed
      atmospheric fields used to force the surface mass balance model show
      that high winter precipitations are mainly confined to the western
      slope of the Coast Mountains, except in the centre of the modelling
      domain where they also occur further inland than along other east-west
      transects (Fig.~\ref{fig:atm}). In fact, along most of the
      north-western coast of North America, coastal mountain ranges form a
      pronounced topographic barrier for westerly winds, capturing
      atmospheric moisture in the form of orographic precipitation, and
      resulting in arid interior lowlands. However, near the centre of our
      modelling domain, this barrier is less pronounced than elsewhere,
      allowing westerly winds to carry moisture further inland, until it is
      captured by the extensive Skeena Mountains in north-central British
      Columbia, thus resulting in a~more widespread distribution of winter
      precipitation (Fig.~\ref{fig:atm}).

@@

      A~correlation is observed between the modelled duration of warm based
      ice cover (Fig.~\ref{fig:hr_geom_warmbase}) and the degree of glacial
      modification of the landscape (mainly in terms of the development of
      deep glacial valleys and troughs). We find evidence for this on the
      slopes of the Coast Mountains, the Columbia and Rocky mountains, the
      Wrangell and Saint Elias mountains, and radiating off the Skeena
      Mountains (Figs.~\ref{fig:hr_geom_duration} and
      \ref{fig:hr_geom_warmbase}; \citealp[Fig.~2]{Kleman.etal.2010}). The
      Skeena Mountains, for example, indeed bear a~strong glacial imprint
      that indicates ice drainage in a~system of distinct glacial troughs
      emanating in a~radial pattern from the centre of the mountain range
      \citep[Fig.~2]{Kleman.etal.2010}. We suggest that persistent ice cover
      (Fig.~\ref{fig:hr_geom_duration}) associated with basal ice
      temperatures at the pressure melting point
      (Figs.~\ref{fig:hr_geom_warmbase} and~\ref{fig:hr_geom_warmfrac})
      explains the large-scale glacial erosional imprint on the landscape. A
      well-developed network of glacial valleys running to the north-west on
      the west slope of the Selwyn and Mackenzie Mountains
      (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
      {Stroeven.etal.2010}) is modelled to have hosted predominantly
      warm-based ice (Fig.~\ref{fig:hr_geom_warmfrac}). However, because it
      is only glaciated for a~short fraction of the last glacial cycle in
      our simulations (Fig.~\ref{fig:hr_geom_duration}), this perhaps
      indicates that the observed landscape pattern originates from multiple
      glacial cycles and witnesses an increased relative importance of the
      Selwyn and Mackenzie mountains ice dispersal centre
      (Fig.~\ref{fig:hr_geom_duration}, SMKM), prior to the Late Pleistocene
      \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.

@@

% -----------------------------------------------------------------------------
\subsection{The last deglaciation}
% -----------------------------------------------------------------------------

@@

      Hence, the two model runs, while similar in overall timing compared to
      the runs with other climate drivers, differ in detail. On the one
      hand, the EPICA run depicts peak glaciation about 2\,\unit{ka} later
      than the GRIP run, in closer agreement with dated maximum extents, and
      shows a faster, uninterrupted deglaciation which yields sporadic ice
      cover more than 1\,\unit{ka} earlier. On the other hand, the GRIP run
      yields a deglaciation in three steps, compatible with marine sediment
      sequences offshore Vancouver Island, where the distribution of
      ice-rafted debris indicates an ice margin retreat from the Georgia
      Strait in two phases that are contemporary with warming oceanic
      temperatures from 17.2 to 16.5 and from 15.5 to
      14.0\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Taylor.etal.2014}.

      Modelled patterns of ice sheet retreat are relatively consistent
      between the two simulations (Figs.~\ref{fig:hr_geom_deglacageshots}
      and.~\ref{fig:hr_geom_deglacage}). The southern sector of the
      modelling domain, including the Puget Lowland, the Coast Mountains,
      the Columbia and and Rocky mountains, and the Interior Plateau of
      British Columbia, becomes completely deglaciated by 10\,\unit{ka},
      whereas a significant ice cover remains over the Skeena, the Selwyn
      and Mackenzie, and the Wrangell and Saint Elias mountains in the
      northern sector of the modelling domain. After 10\,\unit{ka},
      deglaciation continues to proceed across the Liard Lowland with
      a~radial ice margin retreat towards the surrounding mountain ranges,
      consistent with the regional melt water record of the last
      deglaciation \citep{Margold.etal.2013}. Remaining ice continues to
      decay by retreating towards the Selwyn and MacKenzie, and Skeena
      mountains. The last remnants of the Cordilleran ice sheet finally
      disappear from the Skeena Mountains at 6.5\,\unit{ka} (GRIP) and
      6.1\,\unit{ka} (EPICA).

@@

      Because a~general tenant in glacial geomorphology is that the majority
      of landforms (lineations and eskers) are part of the deglacial
      envelope \citep[terminology from][]{Kleman.etal.2006}, having been
      formed close inside the retreating margin of ice sheets
      \citep{Boulton.Clark.1990, Kleman.etal.1997, Kleman.etal.2010}, we
      present maps of basal flow directions immediately preceding
      deglaciation or at the time of cessation of sliding inside a
      cold-based retreating margin (Fig.~\ref{fig:hr_geom_lastflow}). The
      modelled deglacial flow patterns are mostly consistent between the
      GRIP and EPICA simulations. They depict an active ice sheet retreat in
      the peripheral areas, followed by stagnant ice decay in some of the
      interior regions. Several parts of the modelling domain do not
      experience any basal sliding throughout the deglaciation phase
      (Fig.~\ref{fig:hr_geom_lastflow}, hatched areas). This notably
      includes parts of the Interior Plateau in British Columbia, major
      portions of the Alaskan sector of the ice sheet, and a~tortuous ribbon
      running from the Northern Rocky Mountains over the Skeena and Selwyn
      Mountains and into the Mackenzie Mountains.

@@

      On the Interior Plateau of south-central British Columbia, both
      simulations produce a~retreat of the ice margin towards the north-east
      (Fig.~\ref{fig:hr_geom_deglacage}), a~pattern which is validated by
      the geomorphological and stratigraphical record for ancient
      pro-glacial lakes dammed by the retreating ice sheet
      \citep{Perkins.Brennand.2014}. However, the two simulations differ in
      the mode of retreat. The GRIP simulation yields an active retreat with
      basal sliding towards the ice margin to the south, whereas the EPICA
      simulation produces negligible basal sliding on the plateau during
      deglaciation (Fig.~\ref{fig:hr_geom_lastflow}). Yet, the Interior
      Plateau also hosts an impressive set of glacial lineations which
      indicate a~substantial eastwards flow component of the Cordilleran ice
      sheet \citep{Prest.etal.1968, Kleman.etal.2010}. This Interior Plateau
      lineation set could therefore present a~smoking gun for the
      reliability of the presented model results. One explanation for the
      incongruent results could be that the missing feedback mechanisms
      between ice sheet topography and regional climate resulted in a
      modelled ice divide of the LGM ice sheet being too far to the east
      (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
      \citealp{Seguinot.etal.2014}). A~more westerly-located LGM ice divide
      would certainly result in a~different deglacial flow pattern over the
      Interior Plateau. However, a~more westerly-positioned LGM ice divide
      would certainly be associated with a~thinner ice sheet than that
      modelled here. Decreased ice thickness would not promote warm-based
      conditions but, on the contrary, enlarge the region of negligible
      basal sliding (Fig.~\ref{fig:hr_geom_lastflow}). Thus, a~second
      explanation for the incongruent results could be that the Interior
      Plateau lineation system predates deglaciation ice flow, as perhaps
      indicated by some eskers that appear incompatible with these glacial
      lineations \citep[Fig.~9]{Margold.etal.2013a}. Finally, a~third
      explanation could be that local geothermal heat associated with
      volcanic activity on the Interior Plateau could have triggered the
      basal sliding \citep[cf. Greenland ice
      sheet;][]{Fahnestock.etal.2001}.

@@

% =============================================================================
\conclusions
\label{sec:concl}
% =============================================================================

@@

% =============================================================================
\end{document}
\endinput
% =============================================================================

>>>>

\begin{document}

@@

\introduction
\label{sec:intro}

@@

      Although numerical ice sheet modelling has been established as
      a~useful tool to improve our understanding of the Cordilleran ice sheet
      (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991};
      \citealp{Marshall.etal.2000}), the ubiquitously mountainous topography
      of the region has presented two major challenges to its application.
      First, only recent developments in numerical ice sheet models and
      underlying scientific computing tools \citep{Bueler.Brown.2009,
      Balay.etal.2015} have allowed for high-resolution numerical modelling
      of glaciers and ice sheets on mountainous terrain over millennial time
      scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the complex
      topography of the North American Cordillera also induces strong
      geographic variations in temperature and precipitation, thus requiring
      the use of high-resolution climate forcing fields as an input to an
      ice sheet model \citep{Seguinot.etal.2014}. However, evolving climate
      conditions over the last glacial cycle are subject to considerable
      uncertainty and still lie beyond the computational reach of atmosphere
      circulation models.

@@

\section{Model setup}
\label{sec:model}

\subsection{Overview}
\label{sec:overview}

@@

      Basal topography is derived from the ETOPO1 combined topography and
      bathymetry dataset with a~resolution of 1\,arc-min
      \citep{Amante.Eakins.2009}. Sea level is lowered as a~function of time
      based on the Spectral Mapping Project
      \citep[SPECMAP,][]{Imbrie.etal.1989} time scale. Geothermal heat flux
      is applied as a~constant value of 70\,\unit{mW\,m^{-2}} at
      3\,\unit{km} depth (Sect.~\ref{sec:icedyn}). Surface mass balance is
      computed using a~positive degree-day (PDD) model
      (Sect.~\ref{sec:surface}). Climate forcing is provided by a~monthly
      climatology averaged from 1979 to 2000 from the North American
      Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, perturbated
      by time-dependent offsets and lapse-rate temperature corrections
      (Sect.~\ref{sec:atm}).

@@

\subsection{Ice thermodynamics}
\label{sec:icedyn}

      Ice sheet dynamics are typically modelled using a~combination of
      internal deformation and basal sliding. PISM is a~shallow ice sheet
      model, which implies that the balance of stresses is approximated
      based on their predominant components. The Shallow Shelf Approximation
      (SSA) is used as a ``sliding law'' for the Shallow Ice Approximation
      (SIA) by adding velocity solutions of the two approximations
      \citep[Eqns.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}.

@@

      Ice shelf calving is computed using a~double criterion. First,
      a~physically-realistic calving flux is computed based on eigenvalues of
      the horizontal strain rate tensor \citep{Winkelmann.etal.2011,
      Levermann.etal.2012}. This allows floating ice to advance in confined
      embayments, but prevents the formation of extensive ice shelves in the
      open ocean. Second, floating ice thinner than 50\,\unit{m} is
      systematically calved off. A~subgrid scheme by
      \citet{Albrecht.etal.2011} allows for a~continuous migration of the
      calving front. This formulation of calving has been applied to the
      Antarctic ice sheet and has shown to produce a~realistic calving front
      positon for many of the present-day ice shelves
      \citep{Martin.etal.2011}.

\subsection{Surface mass balance}
\label{sec:surface}

@@

\subsection{Climate forcing}
\label{sec:atm}

      Climate forcing driving ice sheet simulations consists of
      a~present-day monthly climatology,
$\{T_{m0}, P_{m0}\}$, where temperatures are modified by offset time series,
${\Delta}T_{TS}$, and lapse-rate corrections, ${\Delta}T_{LR}$:
\begin{align}
    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t)
                    + {\Delta}T_{LR}(t, x, y) \,, \\
    P_m(t, x, y) &= P_{m0}(x, y) \,.
\end{align}

@@

      Finally, lapse-rate corrections, ${\Delta}T_{LR}$, are computed as
      a~function of ice surface elevation, $s$, using the NARR surface
      geopotential height invariant field as a~reference topography,
      $b_{ref}$:
\begin{align}
    {\Delta}T_{LR}(t, x, y) &= -\gamma [s(t, x, y)-b_{ref}] \\
                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{ref}],
\end{align}

@@

\section{Sensitivity to climate forcing time-series}
\label{sec:results}

\subsection{Evolution of ice volume}

@@

\subsection{Extreme configurations}

@@

      In the MIS~3 ice volume minimum reconstructions, a~central ice cap
      persists over the Skeena Mountains (Fig.~\ref{fig:lr_maps}, middle
      panels). Although this ice cap is present in all simulations, its
      dimensions depend sensitively on the choice of the applied
      palaeo-temperature record. Modelled ice volume minima spread over
      a~wide range between 1.69 and 2.88\,\unit{m}\,s.l.e.
      (Table~\ref{tab:extrema}).

@@

\section{Comparison with the geologic record}
\label{sec:discussion}

@@

\subsection{Glacial maxima}

@@

      The exact timing of modelled MIS~2 maximum ice volume depends strongly
      on the choice of applied palaeo-temperature record, which allows for
      a~more in-depth comparison with geological evidence for the timing of
      maximum Cordilleran ice sheet extent. In the Puget Lowland
      (Fig.~\ref{fig:locmap}), the LGM advance of the southern Cordilleran
      ice sheet margin has been constrained by radiocarbon dating on wood
      between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,ka}
      \citep{Porter.Swanson.1998}. These dates are consistent with
      radiocarbon dates from the offshore sedimentary record, which reveals
      an increase of glaciomarine sedimentation between 19.5 and
      16.2\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008,
      Taylor.etal.2014}. Radiocarbon dating of the northern Cordilleran ice
      sheet margin is much less constrained but straddles presented
      constraints from the southern margin. However, cosmogenic exposure
      dating places the timing of maximum CIS extent during the McConnell
      glaciation close to 17\,\unit{\chem{^{10}Be}\,ka}
      \citep{Stroeven.etal.2010, Stroeven.etal.2014}. A~sharp transition in
      the sediment record of the Gulf of Alaska indicates a~retreat of
      regional outlet glaciers onto land at
      14.8\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Davies.etal.2011}.

@@

      During maximum glaciation, both simulations position the main
      meridional ice divide over the western flank of the Rocky Mountains
      (Fig.~\ref{fig:lr_maps}, lower panels; Fig.~\ref{fig:mis2}). This
      result appears to contrast with palaeoglaciological reconstructions
      for central and southern British Columbia with ice divides in a~more
      westerly position, over the western margin of the Interior Plateau
      \citep{Ryder.etal.1991, Stumpf.etal.2000, Kleman.etal.2010,
      Clague.Ward.2011, Margold.etal.2013a}. These indicate that
      a~latitudinal saddle connected ice dispersal centres in the Columbia
      Mountains with the main ice divide \citep{Ryder.etal.1991,
      Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A~latitudinal
      saddle does indeed feature in our modelling results, however, in an
      inverse configuration between the main ice divide over the Columbia
      Mountains and a~secondary divide over the southern Coast Mountains
      (Fig.~\ref{fig:mis2}).

      Such deviation from the geological inferences could reflect the fact
      that our model does not include feedback mechanisms between ice sheet
      topography and the regional climate. Firstly, during the build-up
      phase preceding the LGM, rapid accumulation over the Coast Mountains
      enhanced the topographic barrier formed by these mountain ranges,
      which likely resulted in a~decrease of precipitation and, therefore,
      a~decrease of accumulation in the interior. Secondly, latent warming of
      the moisture-depleted air parcels flowing over this enhanced
      topography could have resulted in an inflow of potentially warmer air
      over the eastern flank of the ice sheet, increasing melt along the
      advancing margin \citep[cf.][]{Langen.etal.2012}. Because these two
      processes, both with a~tendency to limit ice-sheet growth, are absent
      from our model, the eastern margin of the ice sheet and the position
      of the main meridional ice divide are certainly biased towards the
      east in our simulations \citep{Seguinot.etal.2014}.

@@

\subsection{Nucleation centres}

@@

      Perhaps the most striking feature displayed by the distributions of
      modelled ice cover is the persistence of the Skeena Mountains ice cap
      throughout the entire last glacial period (ca.~100--10\,\unit{ka}) and
      its predominance over the other ice-dispersal centres
      (Figs.~\ref{fig:lr_maps} and~\ref{fig:hr_geom_duration}, SM).
      Regardless of the applied forcing, this ice cap appears to survive
      MIS~3 (Fig.~\ref{fig:lr_maps}, middle panels), and serves as
      a~nucleation centre at the onset of the glacial readvance towards the
      LGM (MIS~2). This situation appear similar to the neighbouring
      Laurentide ice sheet, for which the importance of residual ice for the
      glacial history leading up to the LGM has been illustrated by the
      MIS~3 residual ice bodies in northern and eastern Canada as nucleation
      centres for a~much more extensive MIS~2 configuration
      \citep{Kleman.etal.2010}.

      The presence of a~Skeena Mountains ice cap during most of the last
      glacial cycle can be explained by meteorological conditions more
      favourable for ice growth there than elsewhere. In fact, reanalysed
      atmospheric fields used to force the surface mass balance model show
      that high winter precipitations are mainly confined to the western
      slope of the Coast Mountains, except in the centre of the modelling
      domain where they also occur further inland than along other east-west
      transects (Fig.~\ref{fig:atm}). In fact, along most of the
      north-western coast of North America, coastal mountain ranges form
      a~pronounced topographic barrier for westerly winds, capturing
      atmospheric moisture in the form of orographic precipitation, and
      resulting in arid interior lowlands. However, near the centre of our
      modelling domain, this barrier is less pronounced than elsewhere,
      allowing westerly winds to carry moisture further inland, until it is
      captured by the extensive Skeena Mountains in north-central British
      Columbia, thus resulting in a~more widespread distribution of winter
      precipitation (Fig.~\ref{fig:atm}).

@@

      A~correlation is observed between the modelled duration of warm based
      ice cover (Fig.~\ref{fig:hr_geom_warmbase}) and the degree of glacial
      modification of the landscape (mainly in terms of the development of
      deep glacial valleys and troughs). We find evidence for this on the
      slopes of the Coast Mountains, the Columbia and Rocky mountains, the
      Wrangell and Saint Elias mountains, and radiating off the Skeena
      Mountains (Figs.~\ref{fig:hr_geom_duration} and
      \ref{fig:hr_geom_warmbase}; \citealp[Fig.~2]{Kleman.etal.2010}). The
      Skeena Mountains, for example, indeed bear a~strong glacial imprint
      that indicates ice drainage in a~system of distinct glacial troughs
      emanating in a~radial pattern from the centre of the mountain range
      \citep[Fig.~2]{Kleman.etal.2010}. We suggest that persistent ice cover
      (Fig.~\ref{fig:hr_geom_duration}) associated with basal ice
      temperatures at the pressure melting point
      (Figs.~\ref{fig:hr_geom_warmbase} and~\ref{fig:hr_geom_warmfrac})
      explains the large-scale glacial erosional imprint on the landscape.
      A~well-developed network of glacial valleys running to the north-west on
      the west slope of the Selwyn and Mackenzie Mountains
      (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
      {Stroeven.etal.2010}) is modelled to have hosted predominantly
      warm-based ice (Fig.~\ref{fig:hr_geom_warmfrac}). However, because it
      is only glaciated for a~short fraction of the last glacial cycle in
      our simulations (Fig.~\ref{fig:hr_geom_duration}), this perhaps
      indicates that the observed landscape pattern originates from multiple
      glacial cycles and witnesses an increased relative importance of the
      Selwyn and Mackenzie mountains ice dispersal centre
      (Fig.~\ref{fig:hr_geom_duration}, SMKM), prior to the Late Pleistocene
      \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.

@@

\subsection{The last deglaciation}

@@

      Hence, the two model runs, while similar in overall timing compared to
      the runs with other climate drivers, differ in detail. On the one
      hand, the EPICA run depicts peak glaciation about 2\,\unit{ka} later
      than the GRIP run, in closer agreement with dated maximum extents, and
      shows a~faster, uninterrupted deglaciation which yields sporadic ice
      cover more than 1\,\unit{ka} earlier. On the other hand, the GRIP run
      yields a~deglaciation in three steps, compatible with marine sediment
      sequences offshore Vancouver Island, where the distribution of
      ice-rafted debris indicates an ice margin retreat from the Georgia
      Strait in two phases that are contemporary with warming oceanic
      temperatures from 17.2 to 16.5 and from 15.5 to
      14.0\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Taylor.etal.2014}.

      Modelled patterns of ice sheet retreat are relatively consistent
      between the two simulations (Figs.~\ref{fig:hr_geom_deglacageshots}
      and.~\ref{fig:hr_geom_deglacage}). The southern sector of the
      modelling domain, including the Puget Lowland, the Coast Mountains,
      the Columbia and and Rocky mountains, and the Interior Plateau of
      British Columbia, becomes completely deglaciated by 10\,\unit{ka},
      whereas a~significant ice cover remains over the Skeena, the Selwyn
      and Mackenzie, and the Wrangell and Saint Elias mountains in the
      northern sector of the modelling domain. After 10\,\unit{ka},
      deglaciation continues to proceed across the Liard Lowland with
      a~radial ice margin retreat towards the surrounding mountain ranges,
      consistent with the regional melt water record of the last
      deglaciation \citep{Margold.etal.2013}. Remaining ice continues to
      decay by retreating towards the Selwyn and MacKenzie, and Skeena
      mountains. The last remnants of the Cordilleran ice sheet finally
      disappear from the Skeena Mountains at 6.5\,\unit{ka} (GRIP) and
      6.1\,\unit{ka} (EPICA).

@@

      Because a~general tenant in glacial geomorphology is that the majority
      of landforms (lineations and eskers) are part of the deglacial
      envelope \citep[terminology from][]{Kleman.etal.2006}, having been
      formed close inside the retreating margin of ice sheets
      \citep{Boulton.Clark.1990, Kleman.etal.1997, Kleman.etal.2010}, we
      present maps of basal flow directions immediately preceding
      deglaciation or at the time of cessation of sliding inside
      a~cold-based retreating margin (Fig.~\ref{fig:hr_geom_lastflow}). The
      modelled deglacial flow patterns are mostly consistent between the
      GRIP and EPICA simulations. They depict an active ice sheet retreat in
      the peripheral areas, followed by stagnant ice decay in some of the
      interior regions. Several parts of the modelling domain do not
      experience any basal sliding throughout the deglaciation phase
      (Fig.~\ref{fig:hr_geom_lastflow}, hatched areas). This notably
      includes parts of the Interior Plateau in British Columbia, major
      portions of the Alaskan sector of the ice sheet, and a~tortuous ribbon
      running from the Northern Rocky Mountains over the Skeena and Selwyn
      Mountains and into the Mackenzie Mountains.

@@

      On the Interior Plateau of south-central British Columbia, both
      simulations produce a~retreat of the ice margin towards the north-east
      (Fig.~\ref{fig:hr_geom_deglacage}), a~pattern which is validated by
      the geomorphological and stratigraphical record for ancient
      pro-glacial lakes dammed by the retreating ice sheet
      \citep{Perkins.Brennand.2014}. However, the two simulations differ in
      the mode of retreat. The GRIP simulation yields an active retreat with
      basal sliding towards the ice margin to the south, whereas the EPICA
      simulation produces negligible basal sliding on the plateau during
      deglaciation (Fig.~\ref{fig:hr_geom_lastflow}). Yet, the Interior
      Plateau also hosts an impressive set of glacial lineations which
      indicate a~substantial eastwards flow component of the Cordilleran ice
      sheet \citep{Prest.etal.1968, Kleman.etal.2010}. This Interior Plateau
      lineation set could therefore present a~smoking gun for the
      reliability of the presented model results. One explanation for the
      incongruent results could be that the missing feedback mechanisms
      between ice sheet topography and regional climate resulted in
      a~modelled ice divide of the LGM ice sheet being too far to the east
      (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
      \citealp{Seguinot.etal.2014}). A~more westerly-located LGM ice divide
      would certainly result in a~different deglacial flow pattern over the
      Interior Plateau. However, a~more westerly-positioned LGM ice divide
      would certainly be associated with a~thinner ice sheet than that
      modelled here. Decreased ice thickness would not promote warm-based
      conditions but, on the contrary, enlarge the region of negligible
      basal sliding (Fig.~\ref{fig:hr_geom_lastflow}). Thus, a~second
      explanation for the incongruent results could be that the Interior
      Plateau lineation system predates deglaciation ice flow, as perhaps
      indicated by some eskers that appear incompatible with these glacial
      lineations \citep[Fig.~9]{Margold.etal.2013a}. Finally, a~third
      explanation could be that local geothermal heat associated with
      volcanic activity on the Interior Plateau could have triggered the
      basal sliding \citep[cf. Greenland ice
      sheet;][]{Fahnestock.etal.2001}.

@@

\conclusions
\label{sec:concl}

@@

\end{document}
\endinput
